Scalable statistical genetic data analysis in Python

### 2024-01-08 paper.tex
Minimal fix for figures 2 and 3

% Local definitions to systematise tool naming
\newcommand{\sgapi}[1]{\texttt{#1}}
\newcommand{\toolname}[1]{\texttt{#1}}
\newcommand{\sgkit}{\texttt{sgkit}}

@@

\begin{figure}
\begin{tabular}{cc}
\includegraphics[width=6cm]{figures/whole-matrix-compute} &
% \begin{tcolorbox}[width=7cm]
% \begin{minted}[fontsize=\footnotesize]{python}
% def sgkit_afdist(ds, num_bins=10):
%     ds_vs = sgkit.variant_stats(ds).compute()
%     bins = np.linspace(0, 1.0, num_bins + 1)
%     bins[-1] += 0.01
%     af = ds_vs.variant_allele_frequency.values[:, 1]
%     pRA = 2 * af * (1 - af)
%     pAA = af * af
%     hets = ds_vs.variant_n_het.values
%     homs = ds_vs.variant_n_hom_alt.values
%     a = np.bincount(
%         np.digitize(pRA, bins),
%         weights=hets,
%         minlength=num_bins + 1)
%     b = np.bincount(
%         np.digitize(pAA, bins),
%         weights=homs,
%         minlength=num_bins + 1)
%     count = (a + b).astype(int)
%     return pd.DataFrame({
%         "start": bins[:-1],
%         "stop": bins[1:],
%         "prob_dist":
%         count[1:]})

% ds = sgkit.load_dataset(ds_path)
% df = sgkit_afdist(ds)
% \end{minted}
% \end{tcolorbox}
\end{tabular}
\caption{Whole-matrix compute performance with increasing sample size.
(A) Total CPU time required to run \texttt{bcftools +afdist}
and equivalent operations in a single thread for various tools.
Elapsed time is also reported (dotted line).
(B) Relative speedup with 8 threads.
(C) The source code used for the sgkit version of afdist used in these
benchmarks. See Figure~\ref{fig-whole-matrix-compute-supplemental} for
more detailed comparisons with different numbers of threads, and comparing
the effects of HDD vs SSD storage.
\label{fig-whole-matrix-compute}}
\end{figure}

@@

\begin{figure}
\begin{tabular}{cc}
\includegraphics[width=6cm]{figures/subset-matrix-compute} &
% \begin{tcolorbox}[width=7cm]
% \begin{minted}[fontsize=\footnotesize]{python}
% # TODO not sure how best to illustrate this. Trying out various
% # options

% ds = sgkit.load_dataset(ds_path)
% dss = ds.isel(
%     variants=slice(...),
%     samples=(slice(...)))
% df = sgkit_afdist(dss)

% # bcftools
% bcftools view PATH -S ... -r...
% | bcftools +fill-tags | bcftools +af-dist

% # genozip
% genocat PATH -r ... -s ...
% | bcftools +fill-tags | bcftools +af-dist

% # Complex query
% df = pd.read_csv("sample_info.csv",
%     sep="\t").set_index("sample_id")
% dss = ds.isel(
%     samples=df.decade[ds.sample_id.values] >= 1950)
% sgkit_afdist(dss)

% # Corresponding shell
% tail -n+2 sample_info.csv | awk '$2 >= 1950'
%     | cut -f1 > samples.txt
% bcftools view PATH -s samples.txt
%     | bcftools +fill-tags | bcftools +af-dist

% \end{minted}
% \end{tcolorbox}
\end{tabular}
\caption{Whole-matrix compute performance with increasing sample size.
(A) Total CPU time required to run the afdist calculation for
a subset of 10 samples and 10000 variants from the middle of the matrix.
(B) Same for n/2 samples. Genozip did not run in this case for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute.
(C) A simplified version of the source code used for these benchmarks.
\label{fig-subset-matrix-compute}}
\end{figure}

@@

\begin{figure}[ht]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
	\end{center}
	\caption{\label{fig-whole-matrix-compute-supplemental}
    Limitations of vertical scaling on whole-matrix afdist. The elapsed
wall-clock time required to compute afdist on the whole genotype matrix using
different numbers of threads and storage backends (top row), and the relative
speedup obtained from threads as a fraction of optimum speedup (bottom row).
[TODO more detail and link to fig 1. Also add the times for bcftools and
genozip in col 1?]
}
\end{figure}

>>>>

% Additional packages used for code formatting
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
% Local definitions to systematise tool naming
\newcommand{\sgapi}[1]{\texttt{#1}}
\newcommand{\toolname}[1]{\texttt{#1}}
\newcommand{\sgkit}{\texttt{sgkit}}

@@

\begin{figure*}
\begin{tabular}{cc}
\includegraphics[width=6cm]{figures/whole-matrix-compute} &
\begin{tcolorbox}[width=7cm]
\begin{minted}[fontsize=\footnotesize]{python}
def sgkit_afdist(ds, num_bins=10):
    ds_vs = sgkit.variant_stats(ds).compute()
    bins = np.linspace(0, 1.0, num_bins + 1)
    bins[-1] += 0.01
    af = ds_vs.variant_allele_frequency.values[:, 1]
    pRA = 2 * af * (1 - af)
    pAA = af * af
    hets = ds_vs.variant_n_het.values
    homs = ds_vs.variant_n_hom_alt.values
    a = np.bincount(
        np.digitize(pRA, bins),
        weights=hets,
        minlength=num_bins + 1)
    b = np.bincount(
        np.digitize(pAA, bins),
        weights=homs,
        minlength=num_bins + 1)
    count = (a + b).astype(int)
    return pd.DataFrame({
        "start": bins[:-1],
        "stop": bins[1:],
        "prob_dist":
        count[1:]})

ds = sgkit.load_dataset(ds_path)
df = sgkit_afdist(ds)
\end{minted}
\end{tcolorbox}
\end{tabular}
\caption{Whole-matrix compute performance with increasing sample size.
(A) Total CPU time required to run \texttt{bcftools +afdist}
and equivalent operations in a single thread for various tools.
Elapsed time is also reported (dotted line).
(B) Relative speedup with 8 threads.
(C) The source code used for the sgkit version of afdist used in these
benchmarks. See Figure~\ref{fig-whole-matrix-compute-supplemental} for
more detailed comparisons with different numbers of threads, and comparing
the effects of HDD vs SSD storage.
\label{fig-whole-matrix-compute}}
\end{figure*}

@@

\begin{figure*}
\begin{tabular}{cc}
\includegraphics[width=6cm]{figures/subset-matrix-compute} &
\begin{tcolorbox}[width=7cm]
\begin{minted}[fontsize=\footnotesize]{python}
# TODO not sure how best to illustrate this. Trying out various
# options

ds = sgkit.load_dataset(ds_path)
dss = ds.isel(
    variants=slice(...),
    samples=(slice(...)))
df = sgkit_afdist(dss)

# bcftools
bcftools view PATH -S ... -r...
| bcftools +fill-tags | bcftools +af-dist

# genozip
genocat PATH -r ... -s ...
| bcftools +fill-tags | bcftools +af-dist

# Complex query
df = pd.read_csv("sample_info.csv",
    sep="\t").set_index("sample_id")
dss = ds.isel(
    samples=df.decade[ds.sample_id.values] >= 1950)
sgkit_afdist(dss)

# Corresponding shell
tail -n+2 sample_info.csv | awk '$2 >= 1950'
    | cut -f1 > samples.txt
bcftools view PATH -s samples.txt
    | bcftools +fill-tags | bcftools +af-dist

\end{minted}
\end{tcolorbox}
\end{tabular}
\caption{Whole-matrix compute performance with increasing sample size.
(A) Total CPU time required to run the afdist calculation for
a subset of 10 samples and 10000 variants from the middle of the matrix.
(B) Same for n/2 samples. Genozip did not run in this case for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute.
(C) A simplified version of the source code used for these benchmarks.
\label{fig-subset-matrix-compute}}
\end{figure*}

@@

\begin{figure*}[ht]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
	\end{center}
	\caption{\label{fig-whole-matrix-compute-supplemental}
    Limitations of vertical scaling on whole-matrix afdist. The elapsed
wall-clock time required to compute afdist on the whole genotype matrix using
different numbers of threads and storage backends (top row), and the relative
speedup obtained from threads as a fraction of optimum speedup (bottom row).
[TODO more detail and link to fig 1. Also add the times for bcftools and
genozip in col 1?]
}
\end{figure*}
